---
title: 'Lab 8: Logistic Regression'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>

**Jenny Wagner, PhD, MPH**
<br>
Department of Public Health
<br>
California State University, Sacramento
<br>
<br>

# Introduction

In Lab 8, we’ll shift our focus from linear regression (our topic last week) to a new type of model: logistic regression. While linear regression is useful for predicting continuous outcomes (like systolic blood pressure, BMI, or area-level disease incidence rates), logistic regression is used when the *outcome is binary* — that is, when there are only two possible values. In public health research, this often means modeling the presence or absence of a health outcome, such as disease status or event occurrence.

We’ll use the Framingham Heart Study dataset to investigate how baseline characteristics like age, sex, smoking status, BMI, blood pressure, cholesterol, and diabetes relate to the risk of having a heart attack during the study’s follow-up period. Our outcome of interest here is binary: whether or not a participant experienced a heart attack during the follow-up period (yes/no).

Like linear regression, logistic regression allows us to examine associations between predictors and an outcome, but instead of estimating the mean of a continuous outcome, it models the probability of a binary outcome. Because probabilities must fall between 0 and 1, logistic regression uses the logit function (the log of the odds) to transform probabilities into a continuous scale that can be modeled using a linear equation.

There are several important assumptions to keep in mind when using logistic regression:

* The outcome variable is binary.
* Observations are independent of one another.
* There is a linear relationship between the log odds of the outcome and each predictor.
* There is no high multicollinearity among predictors.
* The sample size is sufficiently large for stable estimates 
  * A general rule of thumb is that there should be at least 10 observations per predictor variable. For example, if you plan to use 7 predictor variables in a logistic regression model, you should have a sample size of at least 70 observations.

<br>

# Objectives

After completing Lab 8, you will be able to:

1. Fit a logistic regression model using R/RStudio. 
2. Check assumptions of logistic regression.
3. Exponentiate regression coefficients to obtain odds ratios.
4. Interpret odds ratios.

<br>

# Tasks

For Lab 8, we will continue our exploration of risk factors for cardiovascular disease using the Framingham Heart Study. In this Lab we will focus on a specific outcome: the occurrence of heart attacks during the follow-up period (called `MI_FCHD`). Our goal is to build a logistic regression model using baseline characteristics collected at the start of the study to identify significant predictors of heart attack risk. 

When you are ready to start Lab 8:

1. First create a new R Markdown file using the instructions on the Assignment Guidelines page. Save this `.Rmd` file in a folder dedicated to Lab 8 materials. 
2. Next, download the file called "frmgham2.csv" from Canvas and save it in the same folder as your Lab 8 R Markdown file. Note that this is the original Framingham Heart Study dataset, which we have used in prior Labs.
3. Finally, proceed to read through and carry out each of the tasks detailed below. As usual, you will begin by loading R packages, setting your working directory, and importing the dataset.

<br>

## 1. Install and load R packages

We will use the following packages in this Lab:

```{r, results = FALSE, message = FALSE, warning = FALSE}
# load packages
library(tidyverse)
library(naniar)
library(stats)
library(lmtest)
library(performance) 
```

<br>

## 2. Set your working directory

Set your working directory using the `setwd()` function (see Assignment Guidelines for detailed instructions). 

```
setwd("YOUR FILE PATH HERE")
```

```{r, echo=FALSE}
# set working directory
setwd("~/Google Drive/My Drive/1_CSUS/2024-2025/Spring 2025/PUBH 207B/Website/PUBH-207B-S25")
```

<br>

## 3. Import the dataset into RStudio

Use the `read.csv()` function to import the "frmgham2.csv" dataset. For this to work, this file will need to be saved in the working directory you specified in the above step. 

```{r}
# import dataset
data <- read.csv("frmgham2.csv")
```

<br>

## 4. Prepare the dataset for analysis

As we are starting here with the *original* Framingham Heart Study dataset (and not the cleaned version we have used in a few prior Labs), we will need to follow a few steps to prepare the dataset for analysis. First, we will filter the dataset to contain only the baseline sample. This is because we are going to investigate the association between baseline characteristics and heart attack risk. 

> Note: Although we will be analyzing only the baseline observations, our outcome of interest — whether a participant experienced a heart attack (recorded in the dataset as `MI_FCHD`) applies to the entire follow-up period. This variable is binary, where "1" indicates that a participant experienced a heart attack at some point during follow-up, and "0" indicates they did not.

```{r}
# filter to include only baseline time period
data_sub <- data %>% filter(PERIOD == 1)
```

In analyzing data from prospective cohort studies such as this, the standard approach is to start with a disease-free population and observe the occurrence of new cases over time. As such, we will restrict our analysis to individuals without prevalent heart disease at baseline. We can filter the dataset to include only participants without heart disease at baseline (indicated by the variable `PREVCHD`) as follows:

```{r}
# filter to include only individuals without prevalent heart disease at baseline
data_sub <- data_sub %>% filter(PREVCHD == 0)
```

Next, let's retain only the variables we plan to use in our regression analysis - including our outcome and predictors of interest - so that our dataset is smaller and more manageable. As discussed above, we will investigate sex, age, BMI, systolic blood pressure, total cholesterol level, smoking status, and diabetes status as possible predictors of heart attack risk. We can retain these variables using the `select()` function as follows:

```{r}
# select variables we want to use in regression analysis
data_reg <- data_sub %>% select(RANDID, SEX, AGE, BMI, SYSBP, TOTCHOL, CURSMOKE, DIABETES, MI_FCHD)
```

Before we begin our analysis, let's quickly check for missing data by nesting the `is.na()` function within the `colSums()` function as follows:

```{r}
# count number of missing values for each variable
colSums(is.na(data_reg))
```

From the output, we can see there are 19 missing values for BMI (`BMI`) and 50 missing values for total cholesterol (`TOTCHOL`). Since there are very few missing values relative to our sample size, let's address missingness using complete case analysis (i.e., remove rows with any missing values), as follows:

```{r}
# keep "complete cases" for regression variables
data_reg <- data_reg[complete.cases(data_reg), ] 
```

Finally, as we have done in prior Labs, let's identify variables types and modify them as needed. In the past, we have used the `class()` function to check variable types one at a time. As a quicker alternative, we can use the `str()` command to check the structure of our dataset, including variable types, as follows: 

```{r}
# identify variable type for all variables in the dataset
str(data_reg)
```

Our output contains a list of the variables in our dataset, followed by the variable type, where "num" indicates a variable is a continuous numerical variable and "int" indicates a variable is an integer (i.e., a discrete numerical variable). For our regression analysis, this distinction between continuous numerical variables ("num") and discrete numerical variables ("int") will not matter. However, as we have done many times in the past, we should convert any variables that represent categories (rather than quantities) to factor variables. In our dataset, this includes sex, smoking status, diabetes status, and occurrence of heart attacks during follow-up. We can modify these variables as follows:

```{r}
# convert sex to unordered factor variable
data_reg$SEX <- factor(data_reg$SEX, 
                         levels=c("1", "2"), 
                         ordered=FALSE)
```

```{r}
# convert smoking status to unordered factor variable
data_reg$CURSMOKE <- factor(data_reg$CURSMOKE, 
                         levels=c("0", "1"), 
                         ordered=FALSE)
```

```{r}
# convert diabetes status to unordered factor variable
data_reg$DIABETES <- factor(data_reg$DIABETES, 
                         levels=c("0", "1"), 
                         ordered=FALSE)
```

```{r}
# convert occurrence of heart attack to unordered factor variable
data_reg$MI_FCHD <- factor(data_reg$MI_FCHD, 
                         levels=c("0", "1"), 
                         ordered=FALSE)
```

<br>

## 5. Exploratory data analysis

Before we build our logistic regression model, we should be familiar with the variables we plan to include. First, we should know the distribution of our outcome of interest (i.e., we should know how many people experienced a heart attack during follow-up and how many did not). Understanding the distribution of the outcome provides context for interpreting results and ensures there are enough events to support a stable logistic regression model. We can check the number who experienced a heart attack using the `table()` command as follows:

```{r}
# check how many people in the sample experienced a heart attack during follow-up
table(data_reg$MI_FCHD)
```

From the output, we can see there are 595 participants who experienced a heart attack during follow-up and 3577 participants who did not.

Next, let's check the distribution of our predictor variables of interest, using histograms for numerical variabes and tables for categorical variables, as follows:

```{r}
# check distribution of numerical risk factors
hist(data_reg$AGE, breaks = 30)
hist(data_reg$BMI, breaks = 30)
hist(data_reg$SYSBP, breaks = 30) 
hist(data_reg$TOTCHOL, breaks = 30)
```

```{r}
# check distribution of categorical risk factors
table(data_reg$SEX)
table(data_reg$CURSMOKE)
table(data_reg$DIABETES)
```

The histograms above show that BMI, systolic blood pressure, and total cholesterol are right-skewed. While logistic regression does not require numerical predictor variables to be normally distributed, it does assume that each continuous predictor has a linear relationship with the log-odds of the outcome. Skewed predictors may sometimes violate this assumption. As we move forward with modeling, we’ll need to keep an eye on whether any transformations (e.g., log or square root) or other adjustments are needed to better meet the linearity of the logit assumption. In such cases, achieving more symmetric distributions can help improve model fit. For now, let's move on with fitting our initial logistic regression model.

<br>

## 6. Fit logistic regression model

In this step, we will use the `glm()` function to fit a logistic regression model. Recall that the goal of logistic regression is to examine how a set of predictor variables (in this case, sex, age, BMI, blood pressure, cholesterol, smoking status, and diabetes) are associated with the probability that a participant experienced a heart attack during the follow-up period.

We will specify `family = binomial` because our outcome variable (`MI_FCHD`) is binary, meaning it takes the value "1" if a heart attack occurred and "0" if it did not. The binomial family tells R to use a *logit link function*, which models the log-odds of the outcome as a linear combination of the predictors. In other words, the model estimates how each predictor affects the log-odds of having a heart attack, holding the other variables constant. This allows us to understand the relative contribution of each factor to heart attack risk while adjusting for the others.

```{r}
# fit initial logistic regression model
model1 <- glm(MI_FCHD ~ SEX + AGE + BMI + SYSBP + TOTCHOL + CURSMOKE + DIABETES, 
              family = binomial, data = data_reg)
```

Just as we did with linear regression, we can use the `summary()` function to view the model results, as follows:

```{r}
# generate a summary of the model
summary(model1)
```

The output provides the results of our logistic regression model - which, again, estimates how each predictor is associated with the log-odds of experiencing a heart attack. This may or may not be our "final" model, but for now, let's get familiar with the output:

* **Regression coefficients** (labeled "Estimate")
  * Each Estimate represents the change in the log-odds of having a heart attack for a one-unit increase in that predictor (holding other variables constant).
  * For categorical variables (like SEX2, CURSMOKE1, and DIABETES1), the coefficient compares the listed category to a reference group (more on this later).
* **Statistical significance**
  * The Pr(>|z|) column indicates whether the coefficient is statistically significantly different from 0.
  * Small p-values (e.g., < 0.05) indicate strong evidence that the predictor is associated with the outcome.
  * The asterisks (e.g., ***, **) give a quick visual cue of whether a coefficient is significant.
* **Direction of association**
  * A positive coefficient means that an increase in that predictor increases the log-odds (and therefore the probability) of a heart attack.
  * A negative coefficient means that an increase in that predictor decreases the log-odds (and the probability) of a heart attack.
* **Model fit metrics**
  * Null deviance shows how well a model with *no predictors* fits the data.
  * Residual deviance shows the fit of your current model, where a large reduction from null to residual deviance indicates better model fit.
  * AIC (Akaike Information Criterion) is another measure of model quality, where lower values indicate a better-fitting model when comparing different models.

<br>

## 7. Model diagnostics

Before we interpret and use our model, we first need to run model diagnostics to ensure the assumptions of logistic regression are met. This means checking: (1) whether each numerical predictor variable has a linear relationship with the log-odds of the outcome, and (2) whether there is any concerning multicollinearity between predictor variables.

<br>

### Check linearity assumption

We will use the **Box-Tidwell Test** to check whether each numerical predictor has a linear relationship with the log-odds of the outcome. This test adds an *interaction term* between each numerical variable and its log-transformed version (e.g., AGE * log(AGE)) to the model. If the relationship between the predictor and the log-odds is truly linear, this interaction term should **not** be statistically significant. In a nutshell...

* If the interaction term is not significant → the linearity assumption is likely met for that predictor.
* If the interaction term is significant → the relationship is likely nonlinear, and we may need to use a variable transformation to try to improve linearity.

When using the Box-Tidwell Test, we should first make sure that all values for each numerical predictor variable are greater than 0, as the log of 0 is undefined. To check whether there are any 0 values, you can open the `data_reg` object (click on the object in the Environment window to open it in a new tab), then click on the column name for each numerical variable to quickly sort the values from lowest to highest. If you don't see any 0 values, you can proceed with the test. If needed, you can add a small constant (e.g., 0.01) to any variable that has 0 values.

In our case, there are no 0 values for age, BMI, systolic blood pressure, or total cholesterol, so we will continue on to the first step of the test, which is to create new variables for the log of each numerical predictor, which we can do as follows:

```{r}
# Create new variables for the log of each predictor
data_reg$AGE_log <- log(data_reg$AGE)
data_reg$BMI_log <- log(data_reg$BMI)
data_reg$SYSBP_log <- log(data_reg$SYSBP)
data_reg$TOTCHOL_log <- log(data_reg$TOTCHOL)
```

Next, we will fit a logistic regression model containing all predictors of interest. This time, however, we will also include interaction terms between each numerical predictor variable and its log version, as follows:

```{r}
# Fit a model including interactions between variables and their log
bt_model <- glm(MI_FCHD ~ SEX + AGE + I(AGE*AGE_log) +
                          BMI + I(BMI*BMI_log) +
                          SYSBP + I(SYSBP*SYSBP_log) +
                          TOTCHOL + I(TOTCHOL*TOTCHOL_log) +
                          CURSMOKE + DIABETES,
                data = data_reg, family = binomial)
summary(bt_model)
```

From the model summary, we can see that none of the interaction terms are statistically significant, meaning the linearity assumption is likely met for each numerical predictor, and we won't need to use any variable transformations. Yay! 

<br>

### Check for multicollinearity

Next, we will check for multicollinearity between predictor variables. Recall from our exploration of linear regression that multicollinearity occurs when two or more predictor variables are highly correlated. This can cause problems in estimating regression coefficients. Just as we did in Lab 7, we will use the `check_collinearity()` function to generate a Variance Inflation Factor (VIF) for each variable, as follows:

```{r}
# check for multicollinearity between numerical predictor variables
check_collinearity(model1)
```

From the output, we can see the VIF values (second column from the left) are all close to 1. This means there is no multicollinearity in our original model. Great!

<br>

## 8. Exponentiate regression coefficients

Now that we have verified our original model meets assumptions of logistic regression, we can proceed with interpreting our results. Recall that the coefficients from a logistic regression model represent changes in the log-odds of the outcome for a one-unit increase in each predictor. While this is useful from a modeling standpoint, log-odds can be hard to interpret intuitively. To make our results more meaningful, we can **exponentiate** the coefficients using the `exp()` function. This converts log-odds to **odds ratios**, which are easier to interpret in a public health context. An odds ratio (OR) tells us how the odds of the outcome change with a one-unit increase in the predictor:

* OR > 1 → increased odds (the predictor is a risk factor for the outcome)
* OR < 1 → decreased odds (the predictor is a protective factor for the outcome)
* OR = 1 → no association (there is no association between the predictor and outcome)

We can exponentiate the regression coefficients from our model by nesting the `coef()` function within the `exp()` function and save the result to an R object (called `odds_ratios`), as follows:

```{r}
# exponentiate regression coefficients
odds_ratios <- exp(coef(model1))
```

Let's round the odds ratios to three decimal places and put them in a nice table, as follows:

```{r}
# put odds ratios in a nice table
odds_ratios <- as.data.frame(round(odds_ratios, 3))
print(odds_ratios)
```

In the output, we see two columns - one listing the predictors of interest and the other listing the odds ratio associated with each predictor. In the next section, we'll discuss how to interpret these results.

<br>

## 9. Interpret odds ratios

If you take away only one thing from this Lab, let it be this: how to correctly interpret an odds ratio. Odds ratios are one of the most commonly reported effect measures in public health research involving binary outcomes. They tell us how the odds of an outcome (like having a heart attack) change in relation to a predictor variable:

* For a continuous variable (like age, BMI, systolic blood pressure, or total cholesterol), the odds ratio represents the change in the odds of the outcome associated with a one-unit increase in that variable, holding all other variables constant. 
  * To calculate the percent change in odds, we will use the following formula: (OR−1) × 100 = % change in odds
* For a categorical variable (like sex, smoking status, or diabetes status), the odds ratio compares the odds of the outcome for one category *relative to a reference category*.

For example, we can interpret the odds ratios we obtained in the previous section as follows:

* **SEX2 = 0.303** → Participants coded as SEX=2 (females) have 69.7% lower odds of experiencing a heart attack during follow-up compared to males (the reference group, SEX=1), holding all other variables constant.
* **AGE = 1.023** → For each additional year of age, the odds of experiencing a heart attack increase by 2.3%, holding all other variables constant.
* **BMI = 1.036** → For each one-unit increase in BMI, the odds of having a heart attack increase by 3.6%, holding all other variables constant.
* **SYSBP = 1.016** → For each 1 mmHg increase in systolic blood pressure, the odds of a heart attack increase by 1.6%, holding all other variables constant.
* **TOTCHOL = 1.008** → For each one-unit (mg/dL) increase in total cholesterol, the odds of having a heart attack increase by 0.8%, holding all other variables constant.
* **CURSMOKE1 = 1.365** → Current smokers (CURSMOKE=1) have 36.5% higher odds of experiencing a heart attack compared to non-smokers, holding all other variables constant.
* **DIABETES1 = 2.472** → Participants with diabetes have 2.472 times the odds of having a heart attack compared to those without diabetes — meaning their odds are 147.2% higher, holding all other variables constant.

<br>

# Summary

In Lab 8, you learned how to fit a logistic regression model to explore the relationship between multiple predictors and a binary health outcome. You also checked model assumptions and converted regression coefficients into odds ratios to make results easier to interpret. Most importantly, you practiced interpreting odds ratios, which is a core skill in public health research.

Logistic regression is a powerful tool for identifying and quantifying risk factors for disease. By understanding how variables like age, smoking, or diabetes affect the odds of a health outcome, public health professionals can make informed decisions about prevention strategies, resource allocation, and health policy.

When you are ready, please submit the following to the Lab 8 assignment page on Canvas:

1. An R Markdown document, which has a `.Rmd` extension
2. A knitted `.html` file

Please reach out to me at jenny.wagner@csus.edu if you have any questions. See you in class!